---
title: "Project Notebook"
author: "Keith Engwall"
date: "1/23/2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Objectives
Create a predictive model for identifying patients with Diabetes

See [Capstone Proposal](https://github.com/librariandad/introdatascience/blob/master/capstone_proposal.Rmd) for details.

## Project Dataset
Practice Fusion De-Identified Data Set containing EHR data for approximately 10,000 de-identified patients, including data points for diagnoses, medication, transcript data, and lab observations.  See the [Data Dictionary](https://github.com/librariandad/introdatascience/blob/master/PracticeFusionDataSetDictionary.pdf) for details.

## Project Notes

### Loading data into R
The dataset is contained within an SQLite database file (420.7MB).  To load the data into R requires installation and loading of **RSQLite** and **DBI** R packages.  One of the dependencies is the tibble package.  During the install, I was asked whether to install the binary version (1.3.4) or the source version (1.4.1), which would need compilation.  I wasn't comfortable enough to explore compiling R package code yet, so I went with the 1.3.4 version.

I found this brief [example](http://tiffanytimbers.com/querying-sqlite-databases-from-r/) of how to connect to and query an SQLite database file to be very helpful in getting up and going quickly.

```{r eval=FALSE}
#Connect to SQLite file
con <- dbConnect(SQLite(), dbname="<data.db")

#Define query and store it in my_query
my_query <- dbSendQuery(con, "SELECT name from person_table")
#Fetch data using query and store it in my_data
my_data <- dbFetch(my_query)

#Clear the results cache from my_query
dbClearResult(my_query)

#Perform additional queries

#Disconnect from the database file
dbDisconnect(con)
```
