---
title: "Project Notebook"
author: "Keith Engwall"
date: "1/23/2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Objectives
Create a predictive model for identifying patients with Diabetes

See [Capstone Proposal](https://github.com/librariandad/introdatascience/blob/master/capstone_proposal.Rmd) for details.

## Project Dataset
Practice Fusion De-Identified Data Set containing EHR data for approximately 10,000 de-identified patients, including data points for diagnoses, medication, transcript data, and lab observations.  See the [Data Dictionary](https://github.com/librariandad/introdatascience/blob/master/PracticeFusionDataSetDictionary.pdf) for details.

## General Notes

### Working with SQLite
The dataset is contained within an SQLite database file (420.7MB).  To load the data into R requires installation and loading of **RSQLite** and **DBI** R packages.  One of the dependencies is the tibble package.  During the install, I was asked whether to install the binary version (1.3.4) or the source version (1.4.1), which would need compilation.  I wasn't comfortable enough to explore compiling R package code yet, so I went with the 1.3.4 version.

I found this brief [example](http://tiffanytimbers.com/querying-sqlite-databases-from-r/) of how to connect to and query an SQLite database file to be very helpful in getting up and going quickly.

```{r eval=FALSE}
#Connect to SQLite file
con <- dbConnect(SQLite(), dbname="data.db")

#Define query and store it in my_query
my_query <- dbSendQuery(con, "SELECT name from person_table")
#Fetch data using query and store it in my_data
my_data <- dbFetch(my_query)

#Clear the results cache from my_query
dbClearResult(my_query)

#Perform additional queries

#Disconnect from the database file
dbDisconnect(con)
```

dplyr also has sqlite functions

## Project Notes

### R Packages
The following packages are needed to work on the project.

Note that dbplyr is required to make database connections using dplyr functions.
```{r}
library(tidyr)
library(dplyr)
library(dbplyr)
```

### Connect to database
Note that dbplyr library is required in order for this command to work
```{r}
# Connect to database
my_db <- src_sqlite("compData.db")
```

### Load relevant tables
The tables of interest are:

* *training_diagnosis* contains diagnosis information (we need this in order to identify the patients diagnosed with diabetes)
* *training_allergy* contains allergy information for patient
* *training_medication* contains medication information for patient
* *training_patient* contains gender and year of birth
* *training_smoke* combines patient info with smoking statuses
* *lab tables* contain lab results
    + *training_labResult* contains a record of lab results for a particular transcript
    + *training_labPanel* links training_labResult and training_labObservation
    + *training_labObservation* contains details regarding lab tests

```{r}
diagnosis_tbl <- tbl(my_db, sql("SELECT DiagnosisGuid, PatientGuid, ICD9Code, StartYear, StopYear, Acute FROM training_diagnosis"))
allergy_tbl <- tbl(my_db, sql("SELECT AllergyGuid, PatientGuid, AllergyType, StartYear as AllergyStartYear, ReactionName, SeverityName, MedicationNdcCode FROM training_allergy"))
medication_tbl <- tbl(my_db, sql("SELECT MedicationGuid, PatientGuid, MedicationNdcCode, MedicationName, MedicationStrength, Schedule, DiagnosisGuid FROM training_medication"))
patient_tbl <- tbl(my_db, sql("SELECT PatientGuid, Gender, YearOfBirth FROM training_patient"))
transcript_tbl <- tbl(my_db, sql("SELECT TranscriptGuid, PatientGuid, VisitYear, Height, Weight, BMI, SystolicBP, DiastolicBP, RespiratoryRate, HeartRate, Temperature FROM training_transcript"))
smoke_tbl <- tbl(my_db, sql("SELECT PatientGuid, SmokeEffectiveYear, SmokingStatus_Description, SmokingStatus_NISTCode FROM training_smoke"))
# to get a single table that links the observation data back to the patient,
# join training_labResult, training_labPanel, and training labObservation
lab_tbl <- left_join(left_join(tbl(my_db,sql("SELECT LabResultGuid, PatientGuid FROM training_labResult")),tbl(my_db,sql("SELECT LabResultGuid, LabPanelGuid, PanelName FROM training_labPanel")), by="LabResultGuid"),tbl(my_db,"training_labObservation"), by="LabPanelGuid")
```

### Identify target population  
All Type 1 Diabetes diagnoses have an ICD9Code that starts with 205.  We have a table with all of the diagnoses and one with only the diabetes diagnoses.

```{r}
diabetes_tbl <- tbl(my_db, sql("SELECT DiagnosisGuid, PatientGuid, ICD9Code, StartYear, StopYear, Acute FROM training_diagnosis WHERE ICD9Code LIKE '250%'")) %>% glimpse()
```

### Identify allergy data for target population
```{r warning=FALSE}
# join allergy_tbl to diabetes_tbl to get allergy data for diabetic patients
diabetes_allergy_tbl <- as_tibble(left_join(diabetes_tbl, allergy_tbl, by = c("PatientGuid"))) %>%
  select(PatientGuid, AllergyGuid, AllergyType, ReactionName, SeverityName, AllergyMedicationNdcCode = MedicationNdcCode)

# add has_allergy to diabetes_allergy_tbl to indicate whether diabetic patient has allergies
diabetes_allergy_tbl <- diabetes_allergy_tbl %>%
  mutate(has_allergy = as.integer(!is.na(diabetes_allergy_tbl$AllergyGuid)))

glimpse(diabetes_allergy_tbl)
```

### Identify medication data for target population
Exclude medication specifically given in response to diabetes diagnosis (?)
```{r warning=FALSE}
# join medication_tbl to diabetes_tbl to get medication data for diabetic patients
# filter out medication that is linked to the diabetes diagnosis
diabetes_medication_tbl <- as_tibble(left_join(diabetes_tbl, medication_tbl, by = c("PatientGuid"))) %>% 
  filter(DiagnosisGuid.x != DiagnosisGuid.y) %>% 
  select(PatientGuid, MedicationGuid, MedicationNdcCode)

# add has_meds to diabetes_medication_tbl to indicate whether diabetic patient has medication
diabetes_medication_tbl <- diabetes_medication_tbl %>%
  mutate(has_meds = as.integer(!is.na(diabetes_medication_tbl$MedicationGuid)))

glimpse(diabetes_medication_tbl)

```

### Identify patient information for target population
```{r warning=FALSE}
# join patient_tbl to diabetes_tbl to get gender & age data for diabetic patients
diabetes_patient_tbl <- as_tibble(left_join(diabetes_tbl, patient_tbl, by = c("PatientGuid"))) %>%
  select(PatientGuid, Gender, YearOfBirth) %>% glimpse()
```

### Identify smoking information for target population
Need to parse through results to identify smokers, former smokers, etc.  How should these be grouped?

```{r warning=FALSE}
# join smoke_tbl to diabetes_tbl to get smoking data for diabetic patients
diabetes_smoke_tbl <- as_tibble(left_join(diabetes_tbl, smoke_tbl, by = c("PatientGuid"))) %>%
  select(PatientGuid, SmokeEffectiveYear, SmokingStatus_Description, SmokingStatus_NISTCode) %>%
  arrange(SmokingStatus_NISTCode) %>%
  glimpse()
```

### Identify lab information for target population
```{r warning=FALSE}
# join lab_tbl to diabetes_tbl to get lab results data for diabetic patients
diabetes_lab_tbl <- as_tibble(left_join(diabetes_tbl, lab_tbl, by = c("PatientGuid"))) %>%
  select(PatientGuid, LabObservationGuid, HL7Identifier, HL7Text, HL7CodingSystem, ObservationValue, Units, ReferenceRange, AbnormalFlags, ResultStatus, ObservationYear)

# add has_labs to diabetes_lab_tbl to indicate whether diabetic patient has lab results
diabetes_lab_tbl <- diabetes_lab_tbl %>%
  mutate(has_labs = as.integer(!is.na(diabetes_lab_tbl$LabObservationGuid)))

glimpse(diabetes_lab_tbl)

```

### disconnect from database
This is getting an error.  May have to use a different method to connect
```{r eval=FALSE}
dbDisconnect(my_db)
```