---
title: '3.1 Exercise 1: Basic Data Manipulation'
author: "Keith Engwall"
date: "1/28/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective
Clean a toy data set to get it into a format that can be worked with in R.

## Setup
Load dplyr and tidyr libraries
```{r}
library(dplyr)
library(tidyr)
```
## Load data into R
The data is in a file called refine.csv, which was derived from an Excel file downloaded from [this blog post](http://d3-media.blogspot.nl/2013/11/how-to-refine-your-data.html)

First, verify that refine.csv is in the current directory.
```{r}
dir()
```

Read refine.csv into a data frame.
```{r}
toys <- read.csv("refine_original.csv")
```

## Clean toys data frame

### Clean company column
Start by looking at the levels in toys$company

```{r}
levels(toys$company)
```

There are several versions of each company name, including mixed case, misspellings, etc.  Start by loading the levels into a vector and display it.
```{r}
company_levels <- levels(toys$company)

company_levels
```

There are four company names: Akzo, Philips, Unilever, and Van Houten.  Examining the levels, only the Akzo variations start with a or A, only the Philips variations start with p or P or f, only the Unilever variations start with u or U, and only the Van Houten levels start with v or V.  Load these into separate vectors.

```{r}
akzo_levels <- grep('^[aA]',company_levels, value=TRUE)
philips_levels <- grep('^[pPf]',company_levels, value=TRUE)
unilever_levels <- grep('^[uU]',company_levels, value=TRUE)
vanhouten_levels <- grep('^[vV]',company_levels, value=TRUE)
```

Each of these vectors can be used to collapse the levels in toys$company into the corrected company names

```{r}
levels(toys$company) <- list(Akzo = akzo_levels, Philips = philips_levels, Unilever = unilever_levels, VanHouten = vanhouten_levels)

levels(toys$company)
```

### Separate Product.code...number into two columns

Use tidyr::separate to split this column into two new columns

```{r}
toys <- toys %>% separate(Product.code...number,c("product_code","product_number"),sep = "-")
```

### Create product_category column that replaces product codes with corresponding categories
p = Smartphone
v = TV
x = Laptop
q = Tablet

```{r}
toys <- toys %>% 
  mutate(product_category = 
             replace(product_code,
                   c(
                     grep("p",toys$product_code),
                     grep("v",toys$product_code),
                     grep("x",toys$product_code),
                     grep("q",toys$product_code)
                     ),
                   c(
                     rep("Smartphone",length(which(toys$product_code=="p"))),
                     rep("TV",length(which(toys$product_code=="v"))),
                     rep("Laptop",length(which(toys$product_code=="x"))),
                     rep("Tablet",length(which(toys$product_code=="q"))))
                   )
         )
```

### gather the address components into a single address column

Need to convert these to character vectors first
```{r}
toys$address <- as.character(toys$address)
toys$city <- as.character(toys$city)
toys$country <- as.character(toys$country)
```

Use paste to collapse these into a new column, full_address
```{r}
toys <- toys %>% mutate(full_address = paste(address, city, country, sep = ", "))

toys
```

### create binary columns for each company
create company_[name] columns that contain 1 for rows containing that company and 0 for rows that don't

```{r}
toys <- toys %>% mutate(company_akzo = as.integer(grepl("Akzo",company)))
toys <- toys %>% mutate(company_philips = as.integer(grepl("Philips",company)))
toys <- toys %>% mutate(company_unilever = as.integer(grepl("Unilever",company)))
toys <- toys %>% mutate(company_vanhouten = as.integer(grepl("VanHouten",company)))
```

### create binary columns for each product category
same with product categories

```{r}
toys <- toys %>% mutate(product_smartphone = as.integer(grepl("Smartphone",product_category)))
toys <- toys %>% mutate(product_tv = as.integer(grepl("TV",product_category)))
toys <- toys %>% mutate(product_laptop = as.integer(grepl("Laptop",product_category)))
toys <- toys %>% mutate(product_tablet = as.integer(grepl("Tablet",product_category)))
```

## write to clean csv
```{r}
write_csv(toys, "refine_clean.csv")
```