---
title: '3.1 Exercise 2: Dealing with missing values'
author: "Keith Engwall"
date: "1/28/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup
Load dplyr and tidyr libraries
```{r}
library(dplyr)
library(tidyr)
library(readr)
```

## Load data into R
The data is in a file called titanic3.csv, which was derived from an Excel file downloaded from [this site](http://biostat.mc.vanderbilt.edu/wiki/pub/Main/DataSets/titanic3.xls)

First, verify that refine.csv is in the current directory.
```{r}
dir()
```

Load the data into a tibble using read_csv (this works better than read.csv in that it doesn't automatically store the data as factors)

```{r}
titanic <- read_csv("titanic_original.csv")
```

## Find Missing Values in embarked column
To replace NA values in the embarked column, start by using is.na
```{r}
titanic$embarked[is.na(titanic$embarked)] <- "S"
```

You can confirm the contents of embarked by doing a count on the various values of embarked
```{r}
titanic %>% count(embarked)
```

### Fill in missing ages with mean values
```{r}
titanic$age[is.na(titanic$age)] <- mean(titanic$age, na.rm = TRUE)
```
Other ways to populate the age could have been to create a random age within the range of ages.  The problem with this is that it could skew the mean since a random number doesn't follow the distribution curve.

### Fill in missing boat values with a null string
```{r}
 titanic$boat[is.na(titanic$boat)] <- "NA"
```

### Deal with missing values for cabin number
Leave the missing cabin data and instead create a new column that indicates if the row has a cabin number
```{r}
titanic <- titanic %>% mutate(has_cabin_number = as.integer(!is.na(titanic$cabin)))
```

## Write clean csv
```{r}
write_csv(titanic,"titanic_clean.csv")
```