---
title: "LinearRegression_Homework"
author: "Keith Engwall"
date: "3/31/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
list.files("dataSets")
```

```{r}
# read the states data
states.data <- readRDS("dataSets/states.rds") 
#get labels
states.info <- data.frame(attributes(states.data)[c("names", "var.labels")])
#look at last few labels
tail(states.info, 8)
```


```{r}
# summary of expense and csat columns, all rows
sts.ex.sat <- subset(states.data, select = c("expense", "csat"))
summary(sts.ex.sat)
```

```{r}
# correlation between expense and csat
cor(sts.ex.sat)
```

```{r}
# scatter plot of expense vs csat
plot(sts.ex.sat)
```


```{r}
# Fit our regression model
sat.mod <- lm(csat ~ expense, # regression formula
              data=states.data) # data set
summary(sat.mod)
```

```{r}
summary(lm(csat ~ expense + percent, data = states.data))
```


```{r}
class(sat.mod)
names(sat.mod)
methods(class = class(sat.mod))[1:9]
```


```{r}
confint(sat.mod)
```

```{r}
hist(residuals(sat.mod))
```

```{r}
par(mar = c(4, 4, 2, 2), mfrow = c(1, 2)) #optional
plot(sat.mod, which = c(1, 2)) # "which" argument optional
```

```{r}
# fit another model, adding house and senate as predictors
sat.voting.mod <-  lm(csat ~ expense + house + senate,
                      data = na.omit(states.data))
sat.mod <- update(sat.mod, data=na.omit(states.data))
```

```{r}
# compare using the anova() function
anova(sat.mod, sat.voting.mod)
coef(summary(sat.voting.mod))
```

## Exercise: least squares regression
## ────────────────────────────────────────

##   Use the /states.rds/ data set. Fit a model predicting energy consumed
##   per capita (energy) from the percentage of residents living in
##   metropolitan areas (metro). Be sure to
##   1. Examine/plot the data before fitting the model
##   2. Print and interpret the model `summary'
##   3. `plot' the model to look for deviations from modeling assumptions

##   Select one or more additional predictors to add to your model and
##   repeat steps 1-3. Is this model significantly better than the model
##   with /metro/ as the only predictor?

```{r}
plot(states.data$metro,states.data$energy)
```

```{r}
EnergyModel = lm(energy ~ metro, data = states.data)
summary(EnergyModel)
```


```{r}
par(mar = c(4, 4, 2, 2), mfrow = c(1, 2)) #optional
plot(EnergyModel, which = c(1, 2)) # "which" argument optional
```

```{r}
EnergyModel2 = lm(energy ~ metro + area, data=states.data)
summary(EnergyModel2)
```

```{r}
SSE2 = sum(EnergyModel2$residuals^2)
SSE2
RMSE2 = sqrt(SSE2/nrow(states.data))
RMSE2
mean(states.data$energy, na.rm=TRUE)
```

```{r}
EnergyModel3 = lm(energy ~ metro + area + toxic + green + expense + percent + miles + college, data=states.data)
summary(EnergyModel3)
SSE3 = sum(EnergyModel3$residuals^2)
RMSE3 = sqrt(sum(SSE3/nrow(states.data)))
SSE3
RMSE3
```

```{r}
EnergyModel4 = lm(energy ~ metro + area + toxic + green + college + high + expense + percent + miles + pop, data = states.data)
summary(EnergyModel4)
SSE4 = sum(EnergyModel4$residuals^2)
RMSE4 = sqrt(sum(SSE4/nrow(states.data)))
SSE4
RMSE4
```


```{r}
  #Add the interaction to the model
sat.expense.by.percent <- lm(csat ~ expense*income,
                             data=states.data) 
#Show the results
  coef(summary(sat.expense.by.percent)) # show regression coefficients table
```

```{r}
# make sure R knows region is categorical
str(states.data$region)
states.data$region <- factor(states.data$region)
#Add region to the model
sat.region <- lm(csat ~ region,
                 data=states.data) 
#Show the results
coef(summary(sat.region)) # show regression coefficients table
anova(sat.region) # show ANOVA table
```

```{r}
# print default contrasts
contrasts(states.data$region)
# change the reference group
coef(summary(lm(csat ~ C(region, base=4),
                data=states.data)))
# change the coding scheme
coef(summary(lm(csat ~ C(region, contr.helmert),
                data=states.data)))

```

```{r}
EnergyModel5 = lm(energy ~ metro + area + toxic + green, data=states.data)
summary(EnergyModel5)
SSE5 = sum(EnergyModel5$residuals^2)
RMSE5 = sqrt(sum(SSE5/nrow(states.data)))
SSE5
RMSE5
```

```{r}
EnergyModel6 = lm(energy ~ metro + area + toxic + density*green, data=states.data)
summary(EnergyModel6)
SSE6 = sum(EnergyModel6$residuals^2)
RMSE6 = sqrt(sum(SSE6/nrow(states.data)))
SSE6
RMSE6
```

```{r}
EnergyModel7 = lm(energy ~ metro + area + toxic + density*green + region, data=states.data)
summary(EnergyModel7)
SSE7 = sum(EnergyModel7$residuals^2)
RMSE7 = sqrt(sum(SSE7/nrow(states.data)))
SSE7
RMSE7
```





