---
title: "Project Journal Archived Notes"
author: "Keith Engwall"
date: "2/25/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Archived Notes
The following notes represent notes that are no longer relevant to the project, including supplemental information, discarded methods, etc.

### Working with SQLite
The dataset is contained within an SQLite database file (420.7MB).  To load the data into R requires installation and loading of **RSQLite** and **DBI** R packages.  One of the dependencies is the tibble package.  During the install, I was asked whether to install the binary version (1.3.4) or the source version (1.4.1), which would need compilation.  I wasn't comfortable enough to explore compiling R package code yet, so I went with the 1.3.4 version.

I found this brief [example](http://tiffanytimbers.com/querying-sqlite-databases-from-r/) of how to connect to and query an SQLite database file to be very helpful in getting up and going quickly.

```{r eval=FALSE}
#Connect to SQLite file
con <- dbConnect(SQLite(), dbname="data.db")

#Define query and store it in my_query
my_query <- dbSendQuery(con, "SELECT name from person_table")
#Fetch data using query and store it in my_data
my_data <- dbFetch(my_query)

#Clear the results cache from my_query
dbClearResult(my_query)

#Perform additional queries

#Disconnect from the database file
dbDisconnect(con)
```

dplyr also has sqlite functions

### Connect to database
Note that dbplyr library is required in order for this command to work
```{r}
# Connect to database
my_db <- src_sqlite("compData.db")
```

### Load relevant tables
The tables of interest are:

* *training_diagnosis* contains diagnosis information (we need this in order to identify the patients diagnosed with diabetes)
* *training_allergy* contains allergy information for patient
* *training_medication* contains medication information for patient
* *training_patient* contains gender and year of birth
* *training_smoke* combines patient info with smoking statuses
* *lab tables* contain lab results
    + *training_labResult* contains a record of lab results for a particular transcript
    + *training_labPanel* links training_labResult and training_labObservation
    + *training_labObservation* contains details regarding lab tests

```{r}
diagnosis_tbl <- tbl(my_db, sql("SELECT DiagnosisGuid, PatientGuid, ICD9Code, StartYear, StopYear, Acute FROM training_diagnosis"))
```
```{r}
allergy_tbl <- tbl(my_db, sql("SELECT AllergyGuid, PatientGuid, AllergyType, StartYear as AllergyStartYear, ReactionName, SeverityName, MedicationNdcCode FROM training_allergy"))
medication_tbl <- tbl(my_db, sql("SELECT MedicationGuid, PatientGuid, MedicationNdcCode, MedicationName, MedicationStrength, Schedule, DiagnosisGuid FROM training_medication"))
patient_tbl <- tbl(my_db, sql("SELECT PatientGuid, Gender, YearOfBirth FROM training_patient"))
transcript_tbl <- tbl(my_db, sql("SELECT TranscriptGuid, PatientGuid, VisitYear, Height, Weight, BMI, SystolicBP, DiastolicBP, RespiratoryRate, HeartRate, Temperature FROM training_transcript"))
smoke_tbl <- tbl(my_db, sql("SELECT PatientGuid, SmokeEffectiveYear, SmokingStatus_Description, SmokingStatus_NISTCode FROM training_smoke"))
# to get a single table that links the observation data back to the patient,
# join training_labResult, training_labPanel, and training labObservation
lab_tbl <- left_join(left_join(tbl(my_db,sql("SELECT LabResultGuid, PatientGuid FROM training_labResult")),tbl(my_db,sql("SELECT LabResultGuid, LabPanelGuid, PanelName FROM training_labPanel")), by="LabResultGuid"),tbl(my_db,"training_labObservation"), by="LabPanelGuid")
```

### Identify target population  
All Type 1 Diabetes diagnoses have an ICD9Code that starts with 205.  We have a table with all of the diagnoses and one with only the diabetes diagnoses.

```{r}
diabetes_tbl <- tbl(my_db, sql("SELECT DiagnosisGuid, PatientGuid, ICD9Code, StartYear, StopYear, Acute FROM training_diagnosis WHERE ICD9Code LIKE '250%'")) %>% glimpse()
```

### Identify allergy data for target population
```{r warning=FALSE}
# join allergy_tbl to diabetes_tbl to get allergy data for diabetic patients
diabetes_allergy_tbl <- as_tibble(left_join(diabetes_tbl, allergy_tbl, by = c("PatientGuid"))) %>%
  select(PatientGuid, AllergyGuid, AllergyType, ReactionName, SeverityName, AllergyMedicationNdcCode = MedicationNdcCode)

# add has_allergy to diabetes_allergy_tbl to indicate whether diabetic patient has allergies
diabetes_allergy_tbl <- diabetes_allergy_tbl %>%
  mutate(has_allergy = as.integer(!is.na(diabetes_allergy_tbl$AllergyGuid)))

glimpse(diabetes_allergy_tbl)
```

### Identify medication data for target population
Exclude medication specifically given in response to diabetes diagnosis (?)
```{r warning=FALSE}
# join medication_tbl to diabetes_tbl to get medication data for diabetic patients
# filter out medication that is linked to the diabetes diagnosis
diabetes_medication_tbl <- as_tibble(left_join(diabetes_tbl, medication_tbl, by = c("PatientGuid"))) %>% 
  filter(DiagnosisGuid.x != DiagnosisGuid.y) %>% 
  select(PatientGuid, MedicationGuid, MedicationNdcCode)

# add has_meds to diabetes_medication_tbl to indicate whether diabetic patient has medication
diabetes_medication_tbl <- diabetes_medication_tbl %>%
  mutate(has_meds = as.integer(!is.na(diabetes_medication_tbl$MedicationGuid)))

glimpse(diabetes_medication_tbl)

```

### Identify patient information for target population
```{r warning=FALSE}
# join patient_tbl to diabetes_tbl to get gender & age data for diabetic patients
diabetes_patient_tbl <- as_tibble(left_join(diabetes_tbl, patient_tbl, by = c("PatientGuid"))) %>%
  select(PatientGuid, Gender, YearOfBirth) %>% glimpse()
```

### Identify smoking information for target population
Need to parse through results to identify smokers, former smokers, etc.  How should these be grouped?

```{r warning=FALSE}
# join smoke_tbl to diabetes_tbl to get smoking data for diabetic patients
diabetes_smoke_tbl <- as_tibble(left_join(diabetes_tbl, smoke_tbl, by = c("PatientGuid"))) %>%
  select(PatientGuid, SmokeEffectiveYear, SmokingStatus_Description, SmokingStatus_NISTCode) %>%
  arrange(SmokingStatus_NISTCode) %>%
  glimpse()
```

### Identify lab information for target population
```{r warning=FALSE}
# join lab_tbl to diabetes_tbl to get lab results data for diabetic patients
diabetes_lab_tbl <- as_tibble(left_join(diabetes_tbl, lab_tbl, by = c("PatientGuid"))) %>%
  select(PatientGuid, LabObservationGuid, HL7Identifier, HL7Text, HL7CodingSystem, ObservationValue, Units, ReferenceRange, AbnormalFlags, ResultStatus, ObservationYear)

# add has_labs to diabetes_lab_tbl to indicate whether diabetic patient has lab results
diabetes_lab_tbl <- diabetes_lab_tbl %>%
  mutate(has_labs = as.integer(!is.na(diabetes_lab_tbl$LabObservationGuid)))

glimpse(diabetes_lab_tbl)

```

### disconnect from database
This is getting an error.  May have to use a different method to connect
```{r eval=FALSE}
dbDisconnect(my_db)
```


### no longer using the following to create R data structures
join tables that link on transcript
```{r message = FALSE}
#join diagnosis and transcript
diagJoin <- left_join(diagnosis, transDiag) %>% select(-TranscriptDiagnosisGuid)
diagTran <- left_join(diagJoin, transcript)
diagTran$TranscriptGuid[is.na(diagTran$TranscriptGuid)] <- "DIAG"

#join medication and transcript
medJoin <- left_join(medication, transMed) %>% select(-TranscriptMedicationGuid)
medTran <- left_join(medJoin, transcript)
medTran$TranscriptGuid[is.na(medTran$TranscriptGuid)] <- "MED"

# join allergy and transcript
allerJoin <- left_join(allergy,transAllergy) %>% select(-TranscriptAllergyGuid)
allerTran <- left_join(allerJoin, transcript)
allerTran$TranscriptGuid[is.na(allerTran$TranscriptGuid)] <- "ALLERGY"

# join medTran, diagTran, allerTran, and rest of transcript
allerDiagMedTran <- full_join(full_join(full_join(allerTran,diagTran),medTran),transcript)
```

give non-transcript tables TranscriptGuid column so that they join without causing cartesian products
```{r message=FALSE}
smoke <- smoke %>% mutate(TranscriptGuid = "SMOKE")
immunization <- immunization %>% mutate(TranscriptGuid = "IMMUNIZATION")
```

join remaining tables
```{r message=FALSE}
admts <- full_join(allerDiagMedTran, smoke)
admtsi <- full_join(admts, immunization)
admtsil <- full_join(admtsi, labs)
patient_pop_data <- left_join(patient,admtsil)
```
isolate diabetes patients 
```{r}
diabetes_pop_data <- patient_pop_data %>% filter(dmIndicator == 1)

glimpse(patient_pop_data)

glimpse(diabetes_pop_data)
```




### Age & Gender plot

```{r}
ggplot(patient, aes(x = Gender, y = 2011 - YearOfBirth, col = factor(dmIndicator), 
                    fill = factor(dmIndicator), group = factor(dmIndicator))) +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), position = position_dodge(width = 1))

ggplot(patient, aes(x = Gender, y = 2011 - YearOfBirth, col = factor(dmIndicator), 
                    fill = factor(dmIndicator), group = factor(dmIndicator))) +
geom_boxplot()
```

### BMI plot
```{r}
patientMaster %>%
  filter(BMI < 100) %>%
ggplot(aes(x = 2011 - YearOfBirth, y = BMI, col = factor(dmIndicator))) +
  geom_point()
  #  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), position = position_dodge(width = 1))

```
